---
title: "NEWS"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| message: false
#| warning: false
#| echo: false

{
  use("htmltools", "a")
  use("glue", "glue")
}
```

```{r}
#| echo: false
#| label: funcs

repo <- "https://github.com/tidywf/tidywigits"
vdiff <- function(new, old) {
  stopifnot(utils::compareVersion(new, old) == 1)
  htmltools::a(
    href = glue::glue("{repo}/compare/v{old}...v{new}"),
    glue::glue("v{new} - v{old} diff")
  )
}

pr_or_issue <- function(n, type) {
  x <- c(iss = "issues", pr = "pull")
  stopifnot(is.numeric(n), n > 0, type %in% names(x))
  type2 <- x[type]
  htmltools::a(
    href = glue::glue("{repo}/{type2}/{n}"),
    glue::glue("{type}{n}")
  )
}

iss <- function(n) {
  pr_or_issue(n, "iss")
}

pr <- function(n) {
  pr_or_issue(n, "pr")
}

user <- function(x) {
  a(
    href = glue("https://github.com/{x}"),
    glue("@{x}")
  )
}
```

## v0.1.0 (2026-07-28)

Major refactor. GitHub org migrated from `umccr` to `tidywf`. All 19 tool
configs consolidated from dual `raw.yaml` + `tidy.yaml` files into a single
`schema.yaml` per tool, with parsing simplified across the board (`r pr(189)`).
The nemo API was updated (`nemofy`/`diro` → `run`/`output_dir`, `r pr(200)`).
DVC switched to per-file tracking backed by a public Cloudflare R2 remote
(`r pr(190)`, `r pr(203)`). Docker images now build for both `linux/amd64` and
`linux/arm64` (`r pr(198)`). GHA refactored to use reusable workflows from
`tidywf/actions` (`r pr(181)`).

### Changes

- GitHub org: `umccr` → `tidywf` (URLs, remotes, labels, conda recipe)
- Schema: all tools migrated from `raw.yaml` + `tidy.yaml` to a single
  `schema.yaml`; initially introduced as LinkML (`r pr(174)`) then simplified;
  old config files removed; parsing simplified across all tools (`r pr(189)`);
  simplified version/qc handling (`r pr(205)`); renamed `txt` ftypes to `tsv`,
  added `equal-keyvalue` ftype (`r pr(206)`)
- Flagstats: removed standalone class --- covered by Bamtools
- Bamtools: dropped histogram parser; gene-coverage split into `genecvggenes` +
  `genecvgcvg` tables via `tidy_genecvg_split()` (`r pr(210)`); handle `summary` +
  `wgsmetrics` subtables
- Cobalt: handle `sample` + `buckets` subtables
- Alignments: dropped histogram rows from `markdup` parser
- Linx: fixed germline file handling (`r pr(189)`); added v1.25 test data;
  germline vs somatic outputs now get symmetric `_germline` / `_somatic`
  prefixes via the `refine_files` hook (also tags the somatic side, scoped to
  parsers that actually have a germline counterpart); relies on nemo running the
  hook before disambiguation (`r pr(208)`)
- Purple: germline vs somatic outputs now get symmetric `_germline` / `_somatic`
  prefixes via the `refine_files` hook (`r pr(208)`)
- Sage: added v3.4.4 test data; germline vs somatic outputs (separated by a
  `germline/` / `somatic/` subfolder rather than by basename) now get
  `_germline` / `_somatic` prefixes via `refine_files`, keyed off the parent
  folder, with the variant also injected into `bname` so repeat runs are
  numbered independently per variant (logic in the `refine_by_variant_folder()`
  helper); the gene-coverage file (`.sage.gene.coverage.tsv`) is now parsed by
  Sage's own `genecvg` table rather than the `bamtools` schema, so it is tidied
  as `sage_genecvg` and picks up the germline/somatic tagging, with the
  genes/cvg split shared with Bamtools via `tidy_genecvg_split()` (`r pr(208)`,
  `r pr(210)`)
- Cuppa: handle v1.4 `sampleId` column (fixes `r iss(172)`); use `csv` ftype for
  datacsv; add plotter functions (`r pr(173)`); predsum tidy schema fix
  (`r pr(179)`, `r iss(178)`)
- DVC: switch from directory-level to per-file `.dvc` tracking, backed by public
  Cloudflare R2 (`r pr(190)`, `r pr(203)`); `dvc_download_file()` /
  `dvc_download_all()` download helpers subsequently moved to nemo (`r pr(204)`)
- Wigits: re-add Esvee; use `metapkg`; `WIGITS_TOOLS` changed from character
  vector to named list of R6 classes (`r pr(191)`)
- Vignettes: add `quickstart`, `structure`, `schema_table`, `cicd`, `devnotes`;
  refactor `uml`; remove `schemas_raw` + `schemas_tidy`; consolidate
  installation fragments (`r pr(193)`); add `Output Naming` vignette extending
  nemo's shared template with the Linx/Purple/Sage special cases (`r pr(208)`);
  add Panache format vignettes (`r pr(207)`)
- GHA: use reusable workflows for conda/docker/pkgdown (`r pr(181)`); add
  version-bumping workflow (`r pr(183)`, `r pr(185)`); fix permissions
  (`r pr(182)`); refactor deploy workflow (`r pr(175)`, `r pr(176)`); add
  multi-arch Docker support (`r pr(198)`)
- Move shiny and website to separate repos (`r pr(171)`)
- Pre-commit: replace `lorenzwalthert/precommit` with `posit-dev/air-pre-commit`
  (`air-format`); add `CLAUDE.md` + `new-tool` skill (`r pr(180)`)

### nemo API changes (`r pr(200)`)

- `nemofy()` / `diro` → `run()` / `output_dir` (two-step: first
  `wrangle`/`out_dir`, then `run`/`output_dir`)
- `out_dir` → `output_dir`; `pfix_include` → `prefix_include`; output column
  `input_pfix` → `input_prefix`; CLI flag `--out_dir` → `--output_dir`
- Config/Tool renames: `raw_schemas_all` → `schemas_raw`; `tidy_schemas_all` →
  `schemas_tidy`; `get_tidy_schema` → `get_schema_tidy`; `get_raw_schema` →
  `get_schema_raw`
- All subclasses declare `cloneable = FALSE`
- `.parse_file*()`/`.tidy_file()` public dot methods moved to private in nemo
- `self$get_schema_*()` shortcuts removed from Tool; use
  `self$config$get_schema_*()` directly
- Linx/Purple: germline/somatic prefix logic moved to `initialize()` (later
  reworked into the `refine_files` hook, `r pr(208)`)

### Contributors

- `r user("reisingerf")`: multi-arch Docker support (`r pr(198)`)

## v0.0.7 (2026-02-09)

- cuppa: handle rna predsum (`r pr(170)`, `r iss(169)`)
- s3sync: add isofox + alignments md.metrics

## v0.0.6 (2026-02-04)

- Add Neo support (`r pr(152)`, `r iss(144)`)
- Add Peach support (`r pr(153)`, `r iss(145)`)
- Add Cider support (`r pr(155)`, `r iss(147)`)
- Add Teal support (`r pr(154)`, `r iss(146)`)
- Purple: handle v4.1 (`r pr(150)`, `r iss(141)`)
- Bamtools: support gene/exon coverage files (`r pr(151)`, `r iss(149)`)
- Quarto Website setup for tidywigits outputs (`r pr(157)`, `r pr(160)`,
  `r iss(156)`)
- Add optional redux prefix pattern for Alignments dupfreq (`r iss(161)`,
  `r pr(164)`)
- Add AWS S3 sync wrapper (`r pr(165)`)
- Use renamed `input_id`/`input_pfix`/`diro` for `nemofy` (`r pr(166)`)
- Metadata: override Workflow parent `get_metadata` method for pkg spec
  (`r iss(167)`, `r pr(168)`)

## v0.0.5 (2025-09-07)

- `r vdiff("0.0.5", "0.0.4")`

Major refactor, moving core functionality to
[nemo](https://github.com/tidywf/nemo)

- move Config, Tool, Workflow classes to nemo
- move utils to nemo
- CLI: use nemo wrapper
- use conda env for pkgdown

## v0.0.4 (2025-08-19)

- `r vdiff("0.0.4", "0.0.3")`

Major documentation update - see `r pr(138)` for details.

- test data update
- full README re-write
- pkgdown vignette re-org
- add vignettes for raw/tidy schemas, uml diagram
- add `pkgdown/extra.scss` for CSS customisation
- add logo

## v0.0.3 (2025-08-04)

- `r vdiff("0.0.3", "0.0.2")`

Fixed bug where `normalizePath` was getting called with NULL output directory in
the db format case (`r pr(133)`).

## v0.0.2 (2025-07-14)

- `r vdiff("0.0.2", "0.0.1")`

Mostly Shiny, Conda, Docker, pkgdown and GitHub Actions support.

- Add GitHub Actions for deployment of the following (`r pr(130)`, `r iss(9)`):
  - Conda: add recipe and rattler-builder (`r pr(124)`, `r iss(4)`)
  - Docker: add Dockerfile (`r pr(130)`)
  - Shiny: add summary app (`r pr(122)`)
  - CLI: Add listing and tidy include/exclude support (`r pr(116)`)
  - Add pkgdown support (`r pr(125)`, `r iss(37)`)
  - Add DVC support with some purple test data (`r pr(126)`, `r iss(59)`)
- Optimise file listing (`r pr(131)`, `r iss(127)`)
  - Remove `File` class

## v0.0.1 (2025-06-19)

Initial release of tidywigits.

- R pkg skeleton (`r pr(6)`)
- Add `Config`, `File` and `Tool` classes (`r pr(12)`)
- Add `Workflow` class (`r pr(99)`, `r iss(97)`)
- Add `Oncoanalyser` class (`r pr(52)`, `r pr(101)`, `r iss(39)`)
- Add `bump-my-version` (`r pr(10)`, `r iss(3)`)
- Add `pre-commit` hooks (`r pr(8)`, `r iss(2)`)
- Add `Makefile` and `air.toml` (`r pr(7)`)
- Support for main outputs from the following WiGiTS tools:
  - Alignments (`r pr(28)`, `r pr(79)`, `r iss(17)`, `r iss(77)`)
  - Amber (`r pr(14)`, `r iss(13)`)
  - Bamtools (`r pr(44)`, `r pr(66)`, `r iss(62)`, `r iss(16)`)
  - Chord (`r pr(29)`, `r pr(38)`, `r iss(18)`, `r iss(38)`)
  - Cobalt (`r pr(33)`, `r iss(15)`)
  - Cuppa (`r pr(30)`, `r pr(68)`, `r iss(19)`, `r iss(67)`)
  - Esvee (`r pr(87)`, `r iss(61)`)
  - Flagstats (`r pr(45)`, `r pr(46)`, `r iss(20)`)
  - Isofox (`r pr(80)`, `r iss(76)`)
  - Lilac (`r pr(31)`, `r iss(21)`)
  - Linx (`r pr(50)`, `r pr(89)`, `r iss(22)`, `r iss(88)`)
  - Purple (`r pr(36)`, `r pr(51)`, `r iss(23)`, `r iss(49)`)
  - Sage (`r pr(48)`, `r pr(72)`, `r pr(73)`, `r iss(24)`, `r iss(71)`,
    `r iss(57)`)
  - Sigs (`r pr(47)`, `r iss(25)`)
  - Virusbreakend (`r pr(42)`, `r iss(26)`)
  - Virusinterpreter (`r pr(40)`, `r pr(65)`, `r iss(27)`, `r iss(41)`)
- Schema:
  - versioning (`r pr(64)`, `r iss(43)`)
  - guesser (`r pr(69)`, `r iss(63)`)
- Vignettes:
  - Setup (`r pr(92)`, `r iss(60)`)
- Tool:
  - write functionality (`r pr(98)`, `r iss(95)`)
  - optionally output raw parsed tibbles (`r pr(94)`, `r iss(91)`)
- UML diagram (`r pr(107)`, `r iss(106)`)
- DB schema (`r pr(111)`, `r iss(110)`)
- CLI support (`r pr(114)`)
