---
title: "NEWS"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: pkgs
#| message: false
#| warning: false
#| echo: false
{
  use("htmltools", "a")
  use("glue", "glue")
}
```

```{r echo=FALSE}
repo <- "https://github.com/tidywf/tidywigits"
vdiff <- function(new, old) {
  stopifnot(utils::compareVersion(new, old) == 1)
  htmltools::a(
    href = glue::glue("{repo}/compare/v{old}...v{new}"),
    glue::glue("v{new} - v{old} diff")
  )
}

pr_or_issue <- function(n, type) {
  x <- c(iss = "issues", pr = "pull")
  stopifnot(is.numeric(n), n > 0, type %in% names(x))
  type2 <- x[type]
  htmltools::a(
    href = glue::glue("{repo}/{type2}/{n}"),
    glue::glue("{type}{n}")
  )
}

iss <- function(n) {
  pr_or_issue(n, "iss")
}

pr <- function(n) {
  pr_or_issue(n, "pr")
}

user <- function(x) {
  a(
    href = glue("https://github.com/{x}"),
    glue("@{x}")
  )
}
```

## 0.1.0 (2026-06-04)

### Changes

- Change GitHub org from `umccr` to `tidywf`
- Schema refactor: all tools migrated from old nemo `raw.yaml` + `tidy.yaml`
  configs to a single `schema.yaml` (initially LinkML via `r pr(174)`, then moved
  away from LinkML since it feels overkill). Old config files removed. Parsing
  simplified across all tools. (`r pr(189)`)
- Flagstats: remove standalone class (covered by Bamtools)
- Bamtools: drop histogram parser; genecoverage returned as single table; handle summary + wgsmetrics subtables
- Cobalt: handle sample + buckets subtables
- Alignments: drop histogram from markdup parser
- Linx: fix germline file handling; add v1.25 test data
- Sage: use bamtools genecoverage parser; add v3.4.4 test data
- Cuppa: handle v1.4 `sampleId` column (fixes `r iss(172)`); use csv ftype for datacsv; add plotter functions (`r pr(173)`); predsum tidy schema fix (`r pr(179)`, `r iss(178)`)
- DVC: switch from directory-level to per-file tracking (`r pr(190)`)
- Wigits: re-add Esvee; use metapkg; fall back to `super$get_metadata`; listify `WIGITS_TOOLS` (`r pr(191)`)
- Vignettes: add `structure`, `schema_table`, `cicd`; refactor `uml`; remove `schemas_raw` + `schemas_tidy`; consolidate installation fragments (`r pr(193)`)
- GHA: use reusable workflows for conda/docker/pkgdown (`r pr(181)`); add version-bumping workflow (`r pr(183)`, `r pr(184)`, `r pr(185)`); fix permissions (`r pr(182)`); refactor deploy workflow (`r pr(175)`, `r pr(176)`); add bump.yaml conda env; add multi-arch Docker support (`r pr(198)`)
- Move shiny and website to separate repos (`r pr(171)`)
- Add air formatter to pre-commit hooks; add `CLAUDE.md` + `new-tool` skill (`r pr(180)`)

### nemo API changes

- Renames: `out_dir` => `output_dir`; `pfix_include` => `prefix_include`;
  output column `input_pfix` => `input_prefix`; CLI flag `--out_dir` => `--output_dir`
- Renames (Config/Tool): `raw_schemas_all` => `schemas_raw`,
  `tidy_schemas_all` => `schemas_tidy`, `get_tidy_schema` => `get_schema_tidy`,
  `get_raw_schema` => `get_schema_raw`
- Tool API cleanup (call-site updates): rename `v` to `version` on
  `get_schema_raw`/`get_schema_tidy` calls; rename `name` to `table_name` on
  `.parse_file` calls; replace `self$schemas_raw` with `self$config$schemas_raw`;
  thread `files_tbl` as an explicit argument in `list_files_with_prefix_fn`
  (Bamtools, Chord, Cobalt, Cuppa, Virusbreakend, Purple, Linx)
- `wrangle()` renamed to `run()` on Tool and Workflow; all subclasses declare `cloneable = FALSE`
- `.parse_file*()`/`.tidy_file()` public dot methods moved to private in nemo; updated
  call sites to `private$parse_file*()` (Alignments, Amber, Bamtools, Cobalt, Linx, Purple, Sigs, Virusbreakend)
- `self$get_schema_*()` shortcut wrappers removed from Tool; updated to `self$config$get_schema_*()`
  (Bamtools, Chord, Cobalt, Virusbreakend); `self$config$schemas_raw` field made private,
  updated to `self$config$get_schemas_raw()` (Cuppa)
- Linx/Purple: `list_files()` override no longer feeds `tidy()` in new nemo (which reads
  `private$files` set at init); germline/somatic prefix logic moved to `initialize()`;
  `list_files_with_prefix_fn` helper removed

### Contributors

- `r user("reisingerf")`: multi-arch Docker support (`r pr(198)`)

## v0.0.7 (2026-02-09)

- cuppa: handle rna predsum (`r pr(170)`, `r iss(169)`)
- s3sync: add isofox + alignments md.metrics

## v0.0.6 (2026-02-04)

- Add Neo support (`r pr(152)`, `r iss(144)`)
- Add Peach support (`r pr(153)`, `r iss(145)`)
- Add Cider support (`r pr(155)`, `r iss(147)`)
- Add Teal support (`r pr(154)`, `r iss(146)`)
- Purple: handle v4.1 (`r pr(150)`, `r iss(141)`)
- Bamtools: support gene/exon coverage files (`r pr(151)`, `r iss(149)`)
- Quarto Website setup for tidywigits outputs (`r pr(157)`, `r pr(160)`, `r iss(156)`)
- Add optional redux prefix pattern for Alignments dupfreq (`r iss(161)`, `r pr(164)`)
- Add AWS S3 sync wrapper (`r pr(165)`)
- Use renamed `input_id`/`input_pfix`/`diro` for `nemofy` (`r pr(166)`)
- Metadata: override Workflow parent `get_metadata` method for pkg spec (`r iss(167)`, `r pr(168)`)

## v0.0.5 (2025-09-07)

- `r vdiff("0.0.5", "0.0.4")`

Major refactor, moving core functionality to [nemo](https://github.com/tidywf/nemo)

- move Config, Tool, Workflow classes to nemo
- move utils to nemo
- CLI: use nemo wrapper
- use conda env for pkgdown

## v0.0.4 (2025-08-19)

- `r vdiff("0.0.4", "0.0.3")`

Major documentation update - see `r pr(138)` for details.

- test data update
- full README re-write
- pkgdown vignette re-org
- add vignettes for raw/tidy schemas, uml diagram
- add `pkgdown/extra.scss` for CSS customisation
- add logo

## v0.0.3 (2025-08-04)

- `r vdiff("0.0.3", "0.0.2")`

Fixed bug where `normalizePath` was getting called with NULL output directory in the db format case (`r pr(133)`).

## v0.0.2 (2025-07-14)

- `r vdiff("0.0.2", "0.0.1")`

Mostly Shiny, Conda, Docker, pkgdown and GitHub Actions support.

- Add GitHub Actions for deployment of the following (`r pr(130)`, `r iss(9)`):
  - Conda: add recipe and rattler-builder (`r pr(124)`, `r iss(4)`)
  - Docker: add Dockerfile (`r pr(130)`)
  - Shiny: add summary app (`r pr(122)`)
  - CLI: Add listing and tidy include/exclude support (`r pr(116)`)
  - Add pkgdown support (`r pr(125)`, `r iss(37)`)
  - Add DVC support with some purple test data (`r pr(126)`, `r iss(59)`)
- Optimise file listing (`r pr(131)`, `r iss(127)`)
  - Remove `File` class

## v0.0.1 (2025-06-19)

Initial release of tidywigits.

- R pkg skeleton (`r pr(6)`)
- Add `Config`, `File` and `Tool` classes (`r pr(12)`)
- Add `Workflow` class (`r pr(99)`, `r iss(97)`)
- Add `Oncoanalyser` class (`r pr(52)`, `r pr(101)`, `r iss(39)`)
- Add `bump-my-version` (`r pr(10)`, `r iss(3)`)
- Add `pre-commit` hooks (`r pr(8)`, `r iss(2)`)
- Add `Makefile` and `air.toml` (`r pr(7)`)
- Support for main outputs from the following WiGiTS tools:
  - Alignments (`r pr(28)`, `r pr(79)`, `r iss(17)`, `r iss(77)`)
  - Amber (`r pr(14)`, `r iss(13)`)
  - Bamtools (`r pr(44)`, `r pr(66)`, `r iss(62)`, `r iss(16)`)
  - Chord (`r pr(29)`, `r pr(38)`, `r iss(18)`, `r iss(38)`)
  - Cobalt (`r pr(33)`, `r iss(15)`)
  - Cuppa (`r pr(30)`, `r pr(68)`, `r iss(19)`, `r iss(67)`)
  - Esvee (`r pr(87)`, `r iss(61)`)
  - Flagstats (`r pr(45)`, `r pr(46)`, `r iss(20)`)
  - Isofox (`r pr(80)`, `r iss(76)`)
  - Lilac (`r pr(31)`, `r iss(21)`)
  - Linx (`r pr(50)`, `r pr(89)`, `r iss(22)`, `r iss(88)`)
  - Purple (`r pr(36)`, `r pr(51)`, `r iss(23)`, `r iss(49)`)
  - Sage (`r pr(48)`, `r pr(72)`, `r pr(73)`, `r iss(24)`, `r iss(71)`, `r iss(57)`)
  - Sigs (`r pr(47)`, `r iss(25)`)
  - Virusbreakend (`r pr(42)`, `r iss(26)`)
  - Virusinterpreter (`r pr(40)`, `r pr(65)`, `r iss(27)`, `r iss(41)`)
- Schema: 
  - versioning (`r pr(64)`, `r iss(43)`)
  - guesser (`r pr(69)`, `r iss(63)`)
- Vignettes:
  - Setup (`r pr(92)`, `r iss(60)`)
- Tool:
  - write functionality (`r pr(98)`, `r iss(95)`)
  - optionally output raw parsed tibbles (`r pr(94)`, `r iss(91)`)
- UML diagram (`r pr(107)`, `r iss(106)`)
- DB schema (`r pr(111)`, `r iss(110)`)
- CLI support (`r pr(114)`)
