---
title: "Output Naming"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| include: false

{
  use("nemo")
  use("tidywigits")
  use("fs")
  use("arrow", "read_parquet")
  use("purrr", "map")
  use("dplyr")
  use("glue", "glue")
}
set.seed(1)

# generic template comes from nemo; the special cases below are tidywigits-only
child1 <- system.file(
  "doc-templates/vignettes/_output_naming.qmd",
  package = "nemo"
)
oa <- system.file("extdata/oa", package = "tidywigits")
```

```{r, child=child1}
#| label: child-output-naming
```

## Special cases: semantic prefixes

The disambiguation above (`_2`, `_3`) keeps files from overwriting each other
but tells you nothing about why two inputs collided. Some tools emit germline
and somatic variants of the same table that, after the schema `pattern` is
stripped, reduce to the same prefix and the same `tool_parser`. Left alone they
would collapse to `sample1` / `sample1_2`, which is lossy since you can no
longer tell which file was germline.

A `Tool` subclass (e.g. `Purple`) fixes this by overriding the private
`refine_files` hook. It runs on the base prefix *before* the disambiguation
passes, so a subclass can fold the germline/somatic distinction into the prefix
itself. Once germline and somatic carry different prefixes they no longer
collide, so the generic pipeline never has to invent a `_2`, and any remaining
`_2`/`_3` is a genuine repeat-run marker within a variant.

### Purple: `driver.catalog` germline vs. somatic

- Problem: both `sample1.purple.driver.catalog.germline.tsv` and
  `sample1.purple.driver.catalog.somatic.tsv` are parsed by the `drivercatalog`
  parser, and strip to the prefix `sample1`. There should be a
  `germline`/`somatic` distinguisher string
- Fix: the Purple hook rewrites them by basename into
  `sample1_germline`/`sample1_somatic` *before* the collision check, so they
  never collapse to a lossy `sample1_2` for a single run.

Let us simulate three runs by copying both drivercatalogs (and qc for
comparison) into separate run folders, then initialise one `Purple` object on
the parent:

```{r}
#| label: purple-init

dir_inP <- path(tempdir(), "purple-runs")
for (r in c("run1", "run2", "run3")) {
  dest <- dir_create(path(dir_inP, r))
  file_copy(
    dir_ls(
      path(oa, "purple"),
      regexp = "driver\\.catalog\\.(germline|somatic)\\.tsv$|purple\\.qc$"
    ),
    dest,
    overwrite = TRUE
  )
}
dir_tree(dir_inP)
ppl <- Purple$new(path = dir_inP)
```

Now we tidy the same object. The semantic tag lands first, so any trailing
`_2`/`_3` is a genuine repeat-run marker within `germline` or `somatic`, not a
collision between them. The `prefix_include` option records the prefix as an
`input_prefix` column, and each run/variant lands in its own file:

```{r}
#| label: purple-run

dir_outP <- dir_create(path(tempdir(), "purple-out"))
ppl$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outP,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outP)
```

Stacking the six tidy `drivercatalog` files (top 2 rows) shows that
`input_prefix` distinguishes germline from somatic and one run from the other
after the merge:

```{r}
#| label: purple-merge

dir_ls(dir_outP, regexp = "purple_drivercatalog\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, chrom, gene, cn_min)
# and look at qc too
dir_ls(dir_outP, regexp = "purple_qc\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, qc_status, purity)
```

### Linx: germline vs. somatic annotations

- Problem: several Linx tables have an optional `germline` string in their
  basename (e.g. `linx.germline.breakend.tsv` vs. `linx.breakend.tsv`), so
  `sample1.linx.breakend.tsv` and `sample1.linx.germline.breakend.tsv` both
  reduce to `sample1`.
- Fix: the Linx hook tags both germline and non-germline files but only for
  parsers that actually have a germline file present, so somatic-only tables
  (e.g. `drivers`, `fusion`, `vis_*`) are left untouched.

Let us simulate three runs by copying the paired germline and somatic tables
(and fusions for comparison) into separate run folders, then initialise one
`Linx` object on the parent:

```{r}
#| label: linx-init

dir_inL <- path(tempdir(), "linx-runs")
patl <- "\\.(breakend|links|svs|fusion)\\.tsv$"
linx_files <- c(
  dir_ls(path(oa, "linx/germline_annotations"), regexp = patl),
  dir_ls(path(oa, "linx/somatic_annotations"), regexp = patl)
)
for (r in c("run1", "run2", "run3")) {
  dest <- dir_create(path(dir_inL, r))
  file_copy(linx_files, dest, overwrite = TRUE)
}
dir_tree(dir_inL)
l <- Linx$new(path = dir_inL)
```

Now we tidy the same object. The paired tables (`breakends`, `links`, `svs`)
split cleanly into `sample1_germline` and `sample1_somatic`, and the
somatic-only fusions only need to use `_2`/`_3`:

```{r}
#| label: linx-run

dir_outL <- dir_create(path(tempdir(), "linx-out"))
l$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outL,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outL)
```

Stacking the six tidy `breakends` files (top 2 rows) shows that `input_prefix`
distinguishes germline from somatic and one run from the other after the merge:

```{r}
#| label: linx-merge

dir_ls(dir_outL, regexp = "linx_breakends\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, gene, undisrupted_cn)
```

### Sage: germline vs. somatic folders

- Problem: Linx and Purple carry the germline/somatic distinction in the
  basename. Older versions of Sage write the same basenames (e.g.
  `sample1.sage.bqr.tsv`) into sibling `germline/` and `somatic/` subfolders, so
  the distinction lives in the parent folder, not the filename.
- Fix: the Sage hook takes into account the parent folder and tags the tidy
  files accordingly.

Let us simulate three runs by copying bqrtsv and genecvg into separate run
folders, then initialise one `Sage` object on the parent:

```{r}
#| label: sage-init

dir_inS <- path(tempdir(), "sage-runs")
for (r in c("run1", "run2", "run3")) {
  for (v in c("germline", "somatic")) {
    dest <- dir_create(path(dir_inS, r, v))
    file_copy(
      dir_ls(
        path(oa, "sage", v),
        regexp = "\\.sage\\.bqr\\.tsv$|\\.sage\\.gene\\.coverage\\.tsv$"
      ),
      dest,
      overwrite = TRUE
    )
  }
}
dir_tree(dir_inS)
s <- Sage$new(path = dir_inS)
```

Now we tidy the same object. Note how the output files are distinguished by
including `germline`/`somatic` into the prefix itself, and the trailing
`_2`/`_3` is a genuine repeat-run marker:

```{r}
#| label: sage-run

dir_outS <- dir_create(path(tempdir(), "sage-out"))
s$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outS,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outS)
```

Stacking the `bqrtsv` files (one random row) shows that `input_prefix`
distinguishes germline from somatic across the runs:

```{r}
#| label: sage-merge

dir_ls(dir_outS, regexp = "sage_bqrtsv\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_sample(n = 1)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, dplyr::everything())
```

### When to reach for the hook

- Prefer schema `pattern`s that already separate variants where you can; the
  generic pipeline then needs no help.
- Use `refine_files()` when a single parser matches multiple input files that
  must stay apart with a meaningful label rather than a positional `_2`.
- The hook can touch any `list_files()` column, but rewriting `prefix` is the
  common case since that is what drives the output filename.
