---
title: "Output Naming"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| include: false

use("nemo")
use("tidywigits")
use("fs")
use("arrow", "read_parquet")
use("purrr", "map")
use("dplyr")
use("glue", "glue")
set.seed(1)

# generic template from nemo; special cases below are tidywigits-only
child1 <- system.file(
  "doc-templates/vignettes/_output_naming.qmd",
  package = "nemo"
)
oa <- system.file("extdata/oa", package = "tidywigits")
```

```{r, child=child1}
#| label: child-output-naming
```

## Special cases: semantic prefixes

Germline/somatic variants of a table can strip to the same prefix (`sample1` /
`sample1_2`). `Linx`, `Purple` and `Sage` override the private `refine_files()`
hook (runs *before* disambiguation) to add `_germline`/`_somatic` to the prefix.
Remaining `_2`/`_3` = repeat runs.

### Purple: `driver.catalog` germline vs. somatic

- Problem: `sample1.purple.driver.catalog.{germline,somatic}.tsv` both go
  through the `drivercatalog` parser and strip to prefix `sample1`.
- Fix: the hook rewrites them by basename to `sample1_germline` /
  `sample1_somatic`.

Three runs, both driver catalogs + qc:

```{r}
#| label: purple-init

dir_inP <- path(tempdir(), "purple-runs")
for (r in c("run1", "run2", "run3")) {
  dest <- dir_create(path(dir_inP, r))
  file_copy(
    dir_ls(
      path(oa, "purple"),
      regexp = "driver\\.catalog\\.(germline|somatic)\\.tsv$|purple\\.qc$"
    ),
    dest,
    overwrite = TRUE
  )
}
dir_tree(dir_inP)
ppl <- Purple$new(path = dir_inP)
```

```{r}
#| label: purple-run

dir_outP <- dir_create(path(tempdir(), "purple-out"))
ppl$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outP,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outP)
```

Stacked (top 2 rows each):

```{r}
#| label: purple-merge

dir_ls(dir_outP, regexp = "purple_drivercatalog\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, chrom, gene, cn_min)
# and look at qc too
dir_ls(dir_outP, regexp = "purple_qc\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, qc_status, purity)
```

### Linx: germline vs. somatic annotations

- Problem: `sample1.linx.breakend.tsv` and `sample1.linx.germline.breakend.tsv`
  both reduce to `sample1`.
- Fix: the hook tags both sides, but only for parsers with a germline file
  present; somatic-only tables (`drivers`, `fusion`, `vis_*`) are untouched.

Three runs, paired tables + fusions:

```{r}
#| label: linx-init

dir_inL <- path(tempdir(), "linx-runs")
patl <- "\\.(breakend|links|svs|fusion)\\.tsv$"
linx_files <- c(
  dir_ls(path(oa, "linx/germline_annotations"), regexp = patl),
  dir_ls(path(oa, "linx/somatic_annotations"), regexp = patl)
)
for (r in c("run1", "run2", "run3")) {
  dest <- dir_create(path(dir_inL, r))
  file_copy(linx_files, dest, overwrite = TRUE)
}
dir_tree(dir_inL)
l <- Linx$new(path = dir_inL)
```

Paired tables split into `sample1_germline`/`sample1_somatic`; fusions get
`_2`/`_3`:

```{r}
#| label: linx-run

dir_outL <- dir_create(path(tempdir(), "linx-out"))
l$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outL,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outL)
```

Stacked `breakends` (top 2 rows each):

```{r}
#| label: linx-merge

dir_ls(dir_outL, regexp = "linx_breakends\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, gene, undisrupted_cn)
```

### Sage: germline vs. somatic folders

- Problem: older Sage versions write identical basenames (e.g.
  `sample1.sage.bqr.tsv`) into sibling `germline/` and `somatic/` folders.
- Fix: the hook tags by parent folder (`refine_by_variant_folder()`).

Three runs, `bqr` + `gene.coverage`:

```{r}
#| label: sage-init

dir_inS <- path(tempdir(), "sage-runs")
for (r in c("run1", "run2", "run3")) {
  for (v in c("germline", "somatic")) {
    dest <- dir_create(path(dir_inS, r, v))
    file_copy(
      dir_ls(
        path(oa, "sage", v),
        regexp = "\\.sage\\.bqr\\.tsv$|\\.sage\\.gene\\.coverage\\.tsv$"
      ),
      dest,
      overwrite = TRUE
    )
  }
}
dir_tree(dir_inS)
s <- Sage$new(path = dir_inS)
```

```{r}
#| label: sage-run

dir_outS <- dir_create(path(tempdir(), "sage-out"))
s$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outS,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outS)
```

Stacked `bqrtsv` (one random row each):

```{r}
#| label: sage-merge

dir_ls(dir_outS, regexp = "sage_bqrtsv\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_sample(n = 1)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, dplyr::everything())
```

### When to reach for the hook

- Prefer schema `pattern`s that already separate variants
- Use `refine_files()` when one parser matches files needing a real label, not
  `_2`
- Hook can edit any `list_files()` column; usually `prefix`
