---
title: "Output Naming"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| include: false

{
  use("nemo")
  use("tidywigits")
  use("fs")
  use("arrow", "read_parquet")
  use("purrr", "map")
  use("dplyr")
  use("glue", "glue")
}

# generic template comes from nemo; the special cases below are tidywigits-only
child1 <- system.file(
  "doc-templates/vignettes/_output_naming.qmd",
  package = "nemo"
)
oa <- system.file("extdata/oa", package = "tidywigits")
```

```{r, child=child1}
#| label: child-output-naming
```

## Special cases: semantic prefixes

The disambiguation above (`_2`, `_3`) is purely positional; it keeps files from
overwriting each other but tells you nothing about why two inputs collided. Some
tools emit germline and somatic variants of the same table that, after the
schema `pattern` is stripped, reduce to the same prefix and the same
`tool_parser`. Left alone they would collapse to `sample1` / `sample1_2`, which
is lossy since you can no longer tell which file was germline.

A `Tool` subclass fixes this by overriding the private `refine_files(files)`
hook. It runs on the base prefix *before* the disambiguation passes, so a
subclass can fold the germline/somatic distinction into the prefix itself. Once
germline and somatic carry different prefixes they no longer collide, so the
generic pipeline never has to invent a `_2`, and any remaining `_2`/`_3` is a
genuine repeat-run marker within a variant. `Linx` and `Purple` both use it.

### Purple: `driver.catalog.germline` vs `driver.catalog.somatic`

Purple's `drivercatalog` parser matches both files with a single schema pattern
(`\.purple\.driver\.catalog\.(germline|somatic)\.tsv$`), so both strip to the
prefix `sample1`. The hook rewrites them by basename into `sample1_germline` /
`sample1_somatic` *before* the collision check, so they never collapse to a
lossy `sample1_2`.

Simulate three runs by copying both catalogs into separate run folders, then
initialise one `Purple` on the parent:

```{r}
#| label: purple-init

dirp <- path(tempdir(), "purple-runs")
for (r in c("run1", "run2", "run3")) {
  dest <- dir_create(path(dirp, r))
  file_copy(
    dir_ls(
      path(oa, "purple"),
      regexp = "driver\\.catalog\\.(germline|somatic)\\.tsv$"
    ),
    dest,
    overwrite = TRUE
  )
}
dir_tree(dirp)
ppl <- Purple$new(path = dirp)
```

Now we tidy the same object. The semantic tag lands first, so any trailing
`_2`/`_3` is a genuine repeat-run marker within `germline` or `somatic`, not a
collision between them. The `prefix_include` option records the prefix as an
`input_prefix` column, and each run/variant lands in its own file:

```{r}
#| label: purple-run

dir_outP <- dir_create(path(tempdir(), "purple-out"))
ppl$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outP,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outP)
```

Reading the six `drivercatalog` files and stacking them (top 2 rows),
`input_prefix` keeps the germline/somatic and repeat-run provenance intact once
the filename is gone:

```{r}
#| label: purple-merge

dir_ls(dir_outP, regexp = "purple_drivercatalog\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, chrom, gene, cn_min)
```

### Linx: germline vs. somatic annotations

Several Linx tables have an optional `germline.` segment in their pattern (e.g.
`\.linx\.(germline\.)?breakend\.tsv$`), so `sample1.linx.breakend.tsv` and
`sample1.linx.germline.breakend.tsv` both reduce to `sample1`. Linx tags both
sides, but only for parsers that actually have a germline file present, so
somatic-only tables (`drivers`, `fusion`, `vis_*`) are left untouched.

Let us again simulate three runs by copying the paired germline and somatic
tables into separate folders (including fusions too), then initialise one `Linx`
object on the parent:

```{r}
#| label: linx-init

dirl <- path(tempdir(), "linx-runs")
patl <- "\\.(breakend|links|svs|fusion)\\.tsv$"
linx_files <- c(
  dir_ls(path(oa, "linx/germline_annotations"), regexp = patl),
  dir_ls(path(oa, "linx/somatic_annotations"), regexp = patl)
)
for (r in c("run1", "run2", "run3")) {
  dest <- dir_create(path(dirl, r))
  file_copy(linx_files, dest, overwrite = TRUE)
}
dir_tree(dirl)
l <- Linx$new(path = dirl)
```

Now we tidy the same object. The paired tables (`breakends`, `links`, `svs`)
split cleanly into `sample1_germline` and `sample1_somatic`, and the
somatic-only fusions only need to use `_2`/`3`:

```{r}
#| label: linx-run

dir_outL <- dir_create(path(tempdir(), "linx-out"))
l$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outL,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outL)
```

Stacking the `breakends` files (top 2 rows), similarly `input_prefix` tells
germline from somatic and one run from the next after the merge:

```{r}
#| label: linx-merge

dir_ls(dir_outL, regexp = "linx_breakends\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(
    input_id,
    input_prefix,
    output_id,
    gene,
    undisrupted_cn,
    region_type,
    coding_type
  )
```

### Sage: germline vs. somatic *folders*

Linx and Purple carry the germline/somatic distinction in the *basename*. Sage
is different: it writes the same basenames (e.g. `sample1.sage.bqr.tsv`) into
sibling `germline/` and `somatic/` subfolders, so the distinction lives in the
*parent folder*, not the filename. `Sage$refine_files()` therefore keys off the
folder, folding `germline` / `somatic` into the prefix:

```{r}
#| label: sage-init

dirs <- path(tempdir(), "sage-runs")
for (r in c("run1", "run2", "run3")) {
  for (v in c("germline", "somatic")) {
    dest <- dir_create(path(dirs, r, v))
    file_copy(
      path(oa, "sage", v, "sample1.sage.bqr.tsv"),
      dest,
      overwrite = TRUE
    )
  }
}
dir_tree(dirs)
s <- Sage$new(path = dirs)
```

`list_files()` picks up the folder-derived label, so germline and somatic no
longer collapse into one lossy stream, and each variant is numbered
independently across the three runs:

```{r}
#| label: sage-listfiles

s$list_files() |>
  dplyr::filter(parser == "bqrtsv") |>
  dplyr::select(bname, prefix) |>
  dplyr::arrange(prefix)
```

There is a subtlety in getting the numbering clean. Sage's germline and somatic
files share an *identical* basename (only the folder differs), and nemo's first
disambiguation pass numbers repeat runs `by = "bname"`. Left alone that pass
would treat all six files as one group and interleave the suffixes
(`sample1_germline`, `sample1_germline_3`, `sample1_germline_5`, ...). To avoid
that, the shared `refine_by_variant_folder()` helper also injects the variant
into `bname` (`sample1.germline.sage.bqr.tsv` / `sample1.somatic.sage.bqr.tsv`)
so the two streams are counted separately. The source `path` is left untouched,
so provenance back to the original file is preserved.

```{r}
#| label: sage-run

dir_outS <- dir_create(path(tempdir(), "sage-out"))
s$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outS,
  format = "parquet",
  prefix_include = TRUE
)
dir_tree(dir_outS)
```

Stacking the `bqrtsv` files (top row each), `input_prefix` still separates
germline from somatic across the runs:

```{r}
#| label: sage-merge

dir_ls(dir_outS, regexp = "sage_bqrtsv\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 1)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, alt, ref, context)
```

### When to reach for the hook

- Prefer schema `pattern`s that already separate variants where you can; the
  generic pipeline then needs no help.
- Use `refine_files()` when a single parser legitimately matches multiple input
  files that must stay apart with a meaningful label rather than a positional
  `_2`.
- The hook can touch any `list_files()` column, but rewriting `prefix` is the
  common case since that is what drives the output filename.
