---
title: "Quickstart"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
use("tidywigits")
use("arrow", "read_parquet")
use("fs", "dir_tree")
options(width = 90)
```

{tidywigits} turns raw [WiGiTS/hmftools](https://github.com/hartwigmedical/hmftools) output
directories into consistently structured, versioned, analysis-ready tables. This vignette
walks through the core usage patterns.

## Input

The package ships with a small example dataset under `inst/extdata/oa/`, one subdirectory
per tool:

```{r}
#| label: input_tree
indir <- system.file("extdata/oa", package = "tidywigits")
dir_tree(indir, invert = TRUE, glob = "*.dvc")
```

## Single tool

Each tool has its own R6 class. Construct it with the tool's output directory, then call
`run()` to parse, tidy, and write in one step:

```{r}
#| label: single_tool
#| message: false
outdir <- file.path(tempdir(), "qs_amber")
Amber$new(file.path(indir, "amber"))$run(
  output_dir = outdir,
  format     = "parquet",
  input_id   = "sample1"
)
list.files(outdir, pattern = "\\.parquet$")
```

### File naming

Output files follow the pattern `{prefix}_{tool}_{table}.parquet`, where `prefix` is
derived from the input filenames (here `sample1`). The `metadata.parquet` file is always
written alongside the data files and records input/output paths and package versions.

### Reading a table back

```{r}
#| label: read_back
qc <- read_parquet(file.path(outdir, "sample1_amber_qc.parquet"))
qc
```

## Full WiGiTS run

`Wigits` processes all supported tools in one call. Point it at the parent directory that
contains the per-tool subdirectories:

```{r}
#| label: wigits_run
#| message: false
outdir_w <- file.path(tempdir(), "qs_wigits")
w <- Wigits$new(indir)
w$run(
  output_dir = outdir_w,
  format     = "parquet",
  input_id   = "sample1"
)
list.files(outdir_w, pattern = "\\.parquet$") |> sort()
```

## ID columns

Three optional columns can be prepended to every written table. All are off by default:

| Argument | Column added | Contains |
|---|---|---|
| `input_id = "x"` | `input_id` | sample or run identifier you supply |
| `output_id = "x"` | `output_id` | processing run identifier you supply |
| `prefix_include = TRUE` | `input_prefix` | filename prefix extracted from input files |

These are useful when loading results from multiple samples into the same database table or
combined data frame:

```{r}
#| label: id_cols
#| message: false
outdir_id <- file.path(tempdir(), "qs_amber_id")
Amber$new(file.path(indir, "amber"))$run(
  output_dir     = outdir_id,
  format         = "parquet",
  input_id       = "sample1",
  output_id      = "out1",
  prefix_include = TRUE
)
read_parquet(file.path(outdir_id, "sample1_amber_qc.parquet"))
```

## Metadata

Every `run()` writes a `metadata.parquet` alongside the data files. It records the input
directory, output directory, IDs, and the versions of R packages used:

```{r}
#| label: metadata
read_parquet(file.path(outdir_w, "metadata.parquet")) |> str()
```

## What's next

- **Schema table**: browse every table and column for all 19 tools →
  [schema_table](schema_table.html)
- **Structure**: how schemas, versioning, and the Tool/Workflow class hierarchy work →
  [structure](structure.html) and the
  [nemo docs](https://tidywf.github.io/nemo/articles/structure.html)
- **PostgreSQL**: writing results to a database → [postgresql](postgresql.html)
