---
title: "Quickstart"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
use("tidywigits")
use("arrow", "read_parquet")
use("fs", "dir_tree")
options(width = 90)
```

tidywigits turns raw [WiGiTS/hmftools](https://github.com/hartwigmedical/hmftools) output
directories into consistently structured, versioned, analysis-ready tables. This vignette
walks through the core usage patterns.

## Test data

Example input files are tracked in `inst/extdata/oa/` via [DVC](https://dvc.org/)
and can be downloaded using either of the below options:

- `dvc pull`: requires dvc installation and cloned tidywigits source repo.
- `?nemo::dvc_download_all()`: no additional requirements, uses `download.file()`
  internally.

```r
dir1 <- "extdata/oa/purple" # adjust for whichever tool you want
input_dir <- system.file(dir1, package = "tidywigits")
output_dir <- file.path(tempdir(), "dvc_test")
result <- dvc_download_all(input_dir, output_dir)
```

No credentials are required since the remote is a public Cloudflare R2 bucket.

## Input

Example WiGiTS results:

```{r}
#| label: input_tree
#| code-fold: true
#| code-summary: "View input files"
indir <- system.file("extdata/oa", package = "tidywigits")
dir_tree(indir, invert = TRUE, glob = "*.dvc")
```

## Output - Single Tool

Each tool has its own R6 class. Construct it with the tool's output directory,
then call `run()` to parse, tidy, and write in one step:

```{r}
#| label: single_tool
#| message: false
outdir <- file.path(tempdir(), "qs_ppl")
Purple$new(file.path(indir, "purple"))$run(
  output_dir = outdir,
  format = "parquet",
  input_id = "sample1_id"
)
list.files(outdir, pattern = "\\.parquet$")
```

### File naming

Output files follow the pattern `{prefix}_{tool}_{table}.parquet`, where `prefix` is
derived from the input filenames (here `sample1`). The `metadata.parquet` file is always
written alongside the data files and records input/output paths and package versions.

### Reading a table back

```{r}
#| label: read_back
qc <- read_parquet(file.path(outdir, "sample1_purple_qc.parquet"))
qc |> str()
```

## Output - Full WiGiTS run

`Wigits` processes all supported tools in one call. Point it at the parent directory
that contains the per-tool subdirectories:

```{r}
#| label: wigits_run
#| message: false
outdir_w <- file.path(tempdir(), "qs_wigits")
w <- Wigits$new(indir)
w$run(
  output_dir = outdir_w,
  format = "parquet"
)
list.files(outdir_w, pattern = "\\.parquet$") |> sort() |> str()
```

## ID columns

Three optional columns can be prepended to every written table. All are off by default:

| Argument                | Column added   | Contains                                   |
| ----------------------- | -------------- | ------------------------------------------ |
| `input_id = "x"`        | `input_id`     | sample or run identifier you supply        |
| `output_id = "x"`       | `output_id`    | processing run identifier you supply       |
| `prefix_include = TRUE` | `input_prefix` | filename prefix extracted from input files |

These are useful when loading results from multiple samples into the same database table or
combined data frame:

```{r}
#| label: id_cols
#| message: false
outdir_id <- file.path(tempdir(), "qs_ppl_id")
Purple$new(file.path(indir, "purple"))$run(
  output_dir = outdir_id,
  format = "parquet",
  input_id = "sample1",
  output_id = "out1",
  prefix_include = TRUE
)
read_parquet(file.path(outdir_id, "sample1_purple_qc.parquet"))
```

## Metadata

Every `run()` writes a `metadata.parquet` alongside the data files. It records the input
directory, output directory, IDs, and the versions of R packages used:

```{r}
#| label: metadata
read_parquet(file.path(outdir_w, "metadata.parquet")) |> str()
```

## Other useful articles

- [Schema table](schema_table.html): browse every table and column for all supported WiGiTS tools
- [Structure](structure.html): how schemas, versioning, and the Tool/Workflow class hierarchy work
- [PostgreSQL](postgresql.html): writing results to a database 
