---
title: "Quickstart"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false

use("tidywigits")
use("arrow", "read_parquet")
use("fs", "dir_tree")
options(width = 90)
```

tidywigits turns raw
[WiGiTS/hmftools](https://github.com/hartwigmedical/hmftools) output directories
into versioned, analysis-ready tables.

## Test data

Example inputs live in `inst/extdata/oa/`, tracked via [DVC](https://dvc.org/)
on a public Cloudflare R2 bucket (no credentials). Fetch them with `dvc pull`
from a cloned repo, or from R with `nemo::dvc_download_all()`:

```r
input_dir <- system.file("extdata/oa/purple", package = "tidywigits")
nemo::dvc_download_all(input_dir, file.path(tempdir(), "dvc_test"))
```

## Input

Example WiGiTS results:

```{r}
#| label: input_tree
#| code-fold: true
#| code-summary: "View input files"

indir <- system.file("extdata/oa", package = "tidywigits")
dir_tree(indir, invert = TRUE, glob = "*.dvc")
```

## Output - Single Tool

Each tool has its own R6 class. `run()` parses, tidies and writes in one step:

```{r}
#| label: single_tool
#| message: false

outdir <- file.path(tempdir(), "qs_ppl")
Purple$new(file.path(indir, "purple"))$run(
  output_dir = outdir,
  format = "parquet",
  input_id = "sample1_id"
)
list.files(outdir, pattern = "\\.parquet$")
```

### File naming

Output files follow `{prefix}_{tool}_{table}.parquet`, where `prefix` comes from
the input filenames (here `sample1`). See [Output Naming](output_naming.html)
for collision handling.

### Reading a table back

```{r}
#| label: read_back

qc <- read_parquet(file.path(outdir, "sample1_purple_qc.parquet"))
qc |> str()
```

## Output - Full WiGiTS run

`Wigits` runs all supported tools on a parent directory:

```{r}
#| label: wigits_run
#| message: false

outdir_w <- file.path(tempdir(), "qs_wigits")
w <- Wigits$new(indir)
w$run(
  output_dir = outdir_w,
  format = "parquet"
)
list.files(outdir_w, pattern = "\\.parquet$") |> sort() |> str()
```

## ID columns

Optional columns prepended to every written table (all off by default), useful
when combining samples into one table:

| Argument                | Column added   | Contains                                   |
| ----------------------- | -------------- | ------------------------------------------ |
| `input_id = "x"`        | `input_id`     | sample or run identifier you supply        |
| `output_id = "x"`       | `output_id`    | processing run identifier you supply       |
| `prefix_include = TRUE` | `input_prefix` | filename prefix extracted from input files |

```{r}
#| label: id_cols
#| message: false

outdir_id <- file.path(tempdir(), "qs_ppl_id")
Purple$new(file.path(indir, "purple"))$run(
  output_dir = outdir_id,
  format = "parquet",
  input_id = "sample1",
  output_id = "out1",
  prefix_include = TRUE
)
read_parquet(file.path(outdir_id, "sample1_purple_qc.parquet"))
```

## Metadata

Every `run()` also writes metadata (input/output dirs, IDs, R package versions):
`metadata.parquet` for `Wigits`, `metadata_<tool>.parquet` for a single tool.

```{r}
#| label: metadata

read_parquet(file.path(outdir_w, "metadata.parquet")) |> str()
```

## Other useful articles

- [Schema table](schema_table.html): browse every table and column for all
  supported WiGiTS tools
- [Structure](structure.html): schemas, versioning, and the Tool/Workflow class
  hierarchy (nemo)
- [PostgreSQL](postgresql.html): writing results to a database
